Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-05 application.tex
rough re-write of sec 7 done by mkl

Perturbations are in the optically thin regime,
$l\sim 1/k_x \lesssim\l_\mathrm{rad}$, when 
\begin{align}
  \khat\gtrsim 2.5\times10^3\hat{\kappa}_d\hat{\rho}(z)r_\mathrm{AU}^{-33/14}.  
\end{align}
In the inner disk, e.g. $r\sim 1$AU, only
extremely small-scale perturabtions are in the optically-thin
regime. However, in the outer disk, e.g. $r\sim 10$AU, the transition
occurs at more moderate wavenumbers $\khat\sim 10$.
%, though the
%wavelength is still small compared to $H$.

@@

\subsection{Comparing $\beta$ to $\beta_\mathrm{crit}$ in the MMSN}

Fig. \ref{beta_compare} show examples of
$\beta$ at $r=5$AU and $r=50$AU for a perturbation wavenumber
$\khat=30$.

% In the outer disk  at $r=50$AU, $\beta$ is vertically-uniform and is less
% than $\beta_\mathrm{crit}$.  Accordingly we expect, and find, the VSI
% operates.    
% In the inner disk ($r=5$AU), the perturbation is in the optically
% thick regime near the midplane, but becomes optically-thin away from
% the midplane.  In this case, $\beta < \beta_\mathrm{crit}$
% for $|z|\gtrsim1.25H$. In fact, we also find the VSI in this case, but
% with much reduced growth rates (when scaled by the local rotation), as
% described below.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

In Fig. \ref{mmsn_bcrit_bcool} we compare the above thermal timescale,
Eq. \ref{beta_mmsn_simp},  for several opacity scales, to  
the VSI criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

For disturbances in the optically-thin cooling regime, $\beta$ does
not depend on the height $z$ or perturbation lengthscale $\khat$. This
generally occurs at sufficiently large radii. In this case we may
directly compare $\beta$ to $\beta_\mathrm{crit}$, which was derived
assuming it is vertically-uniform.  For the case with MMSN
opacity ($\hat{\kappa}_d=1$), we  infer that the VSI can operate at
10s of AU.

For disturbances in the optically-thick or diffusive regime, $\beta$
increases with decreasing $\khat$ (larger scales) and decreases with
increasing height, on account of the density-dependence of
$t_\mathrm{diff}$.  This generally occurs at small radii. In this case 
$\beta$ may be smaller or greater than $\beta_\mathrm{crit}$ depending
on $z$. Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in \S\ref{vsi_mmsn_grow}.

@@

For increasing opacity, $\hat{\kappa}_d=1\to 10$,  a comparison
between the middle and lower panels shows that the optically-thick
regions shift outward (where the three curves are distinct); while the
optically-thin regions have lowered cooling times (where the curves
overlap). Thus, instability is expected to be restricted to larger
radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
at small radii for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane at, say $r\sim 1$AU, the VSI would only occur on very small
radial lengthscales $\ll O(10^{-2}H)$. It is, however, unclear if such
small-scale disturbances are dynamically important.

@@

% \\
% \emph{For this figure to belong here (and not in the previous subsection, with the other comparisons of $\beta$ and $\beta_{crit}$) the discussion should be be more focused on the linear analysis.  Again it's better to start by discussing what the figure tells us, than repeating the figure caption. E.g.\ you could use Fig 21 to explain when a height varying $\beta$ is required.  I would also explain (especially if it stays in this subsection) what the growth time is for the r = 5 AU case.  Looks to be $\sim 1000$ orbits, which is remakrable given that $\beta < \beta_{\rm crit}$ outside $z \gtrsim  1.6 H$.?  That kind of detail could be illuminating, as it's nominally surprising given previous evidence that behavior near 5 -- 7 $H$ can affect growth.}

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$. We do not
consider surface modes since they are boundary-dependent and likely
subject to viscous decay (see). 
%  which are 
% boundary-dependent. 
% Disturbances with large $\khat$ are also more likely subject to viscous 
% decay, as discussed above. 
 %\\
%\emph{above sentence too long and complex, could e.g.\ split off that surface modes are neglected and why.}
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the fiducial MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as
perturbations generally become optically thick.

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU (and rapidly increases to $> 10^3P_0$ for
smaller $r$). Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

@@

We thus find Fig. \ref{mmsn_overall} to be roughly consistent with
comparing the midplane thermal relaxation timescales to the critical
thermal timescale. For example, 
Fig. \ref{mmsn_bcrit_bcool} show that $\beta(z=0,\khat=10) \gtrsim
\beta_\mathrm{crit}$ for $r\lesssim 10$AU in the MMSN (left middle
panel); and Fig. \ref{mmsn_overall} 
show that modes with $\khat=10$ are stabilized inside $\sim 8$AU (where
$t_\mathrm{grow}\to\infty$). Our numerical results suggest the VSI
is efficient at radii of few tens of AU in the MMSN, consistent with
estimates made by \citetalias{nelson13}.

>>>>

Perturbations are in the optically thin regime,
$l\sim 1/k_x \lesssim\l_\mathrm{rad}$, when 
\begin{align}
  \khat\gtrsim 2.5\times10^3\hat{\kappa}_d\hat{\rho}(z)r_\mathrm{AU}^{-33/14}.  
\end{align}
In the inner disk, e.g. $r_\mathrm{AU}\sim 1$, only
extremely small-scale perturabtions are in the optically-thin
regime. However, in the outer disk, e.g. $r_\mathrm{AU}\sim 10$, the transition
occurs at more moderate wavenumbers $\khat\sim 10$.
%, though the
%wavelength is still small compared to $H$.

@@

\subsection{Comparing $\beta$ to $\beta_\mathrm{crit}$ in the MMSN} 
In this section we compare the above thermal timescale, Eq. \ref{beta_mmsn_simp}, to 
the VSI criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}
Fig. \ref{beta_compare} show examples of $\beta$ in the fiducial disk
at $r_\mathrm{AU}=5$ and $r_\mathrm{AU}=50$ for a perturbation wavenumber $\khat=30$.  
At $r_\mathrm{AU}=50$, the disturbance is in the optically-thin regime,  
$\beta$ is vertically uniform, as assumed for our analysis, and $\beta
< \beta_\mathrm{crit}\, \forall z$. Accordingly we expect, and find, the VSI
operates.

On the other hand, at $r_\mathrm{AU}=5$, the perturbation is in the optically
thick (thin) regime near (away from) the midplane. In this case,
$\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. Our analytic 
criteria is not valid for vertically non-uniform $\beta$ so we must
turn to numerical solutions of the linear problem. We do this in
\S\ref{vsi_mmsn_grow}. % and find the VSI also operates in this case, but
% with significantly reduced efficiency.

%, as
% described below.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the fiducial MMSN at $r=50$AU
    and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

In Fig. \ref{mmsn_bcrit_bcool}, we compare $\beta$ to
$\beta_\mathrm{crit}$ for $r_\mathrm{AU}\in[1,100]$, as well as for different
opacity scales. Notice that $\beta$ increases for either large or
small radii. For sufficiently large $r$, disturbances are generally
in the optically-thin cooling regime, characterized graphically by the
over-lapping curves. 
%, where $\beta$ is independent of
%$z$ and the perturbation lengthscale $\khat$.
In this case, we may
directly compare $\beta$ to $\beta_\mathrm{crit}$ and infer that the
VSI can operate at 10s of AU in the fiducial MMSN disk (with
$\hat{\kappa}_d=1$).  However, at small radii, $\beta$ depends on both
$z$ and $\khat$, as the disturbance enters the optically-thick limit
at the midplane. This regime is characterized by the distinct curves for each
$\khat$ toward decreasing $r$.

@@

For increasing opacity, $\hat{\kappa}_d=1\to 10$,  a comparison
between the middle and lower panels shows that the optically-thick
regions shift outward; while the
optically-thin regions have lowered cooling times. Thus, instability
is expected to be restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$ and small $r$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane at, say $r_\mathrm{AU}\sim 1$, the VSI would only occur on very small
radial lengthscales $\ll O(10^{-2}H)$. It is, however, unclear if such
small-scale disturbances are dynamically important.

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$. We discard
surface modes for this discussion since they are not well-defined 
in vertically isothermal disk models \citepalias{barker15} and may be 
subject to viscous decay in real disks (see \S\ref{caveats_visc}).

%  which are 
% boundary-dependent. 
% Disturbances with large $\khat$ are also more likely subject to viscous 
% decay, as discussed above. 
 %\\
%\emph{above sentence too long and complex, could e.g.\ split off that surface modes are neglected and why.}
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the fiducial MMSN. We see the VSI operates from $r_\mathrm{AU}\sim 5$ to $r_\mathrm{AU}\sim
50$ with growth timescales $\sim 30$---40 orbits and 
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as
perturbations generally become optically thick.

The fact that we can obtain the VSI in the despite $\beta >
\beta_\mathrm{crit}$ near the midplane (e.g. $\khat = 30$ at
$r_\mathrm{AU} =5$, see Fig. \ref{beta_compare}) suggests  that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r_\mathrm{AU}=50$, but $t_\mathrm{grow}\sim
400$ orbits at $r_\mathrm{AU}=5$ (and rapidly increases to $> 10^3$ orbits for
smaller $r$). Thus, introducing an optically-thick midplane 
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

@@

We thus find Fig. \ref{mmsn_overall} to be roughly consistent with
comparing the midplane thermal relaxation timescales to the critical
thermal timescale. For example, 
Fig. \ref{mmsn_bcrit_bcool} show that $\beta(z=0,\khat=10) \gtrsim
\beta_\mathrm{crit}$ for $r_\mathrm{AU}\lesssim 10$ in the MMSN (left middle
panel); and Fig. \ref{mmsn_overall} 
show that modes with $\khat=10$ are stabilized for $r_\mathrm{AU}\lesssim 8$ (where
$t_\mathrm{grow}\to\infty$). Our numerical results suggest the VSI
is efficient at radii of few tens of AU in the MMSN, consistent with
estimates made by \citetalias{nelson13}.
